Type Flattening Obfuscation

### 2020-02-02 paper/tool.tex
update

\usepackage{listings}
\lstset{basicstyle=\small\ttfamily,showspaces=false,frame=tb}
\lstloadlanguages{C,C++,[x86masm]Assembler}

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the "assumed" original language.
In a statically typed language like C, the compiler does not preserve the high-level
type information in the generated machine code,
% Because most of information about high-level types is lost in compilation,
% the machine code
then the \emph{type reconstruction} (or type recovery) from machine code requires
special techniques, a survey up until 2015 can be referenced in~\cite{caballero_type_2016}, more
recent research includes~\cite{noonan_polymorphic_2016, maier_typeminer_2019, katz_estimating_2016}.

@@

\subsection*{Type flattening obfuscation}
Under our interest, the studies in binary code type reconstruction can be roughly divided into
two directions: one is based on semantics of programs, the other is not. We focus on the former,
and also on methods using static analysis. The research in this vein starts with
Mycroft~\cite{mycroft_type-based_1999}, and improved in~\cite{lee_tie_2011,noonan_polymorphic_2016,elwazeer_scalable_2013}

\section{Compiler for type flattening}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types, \emph{uCc}'s
objective is to make reconstructing C types from machine codes harder.
The compiler is implemented in Rust, the source code and a brief user guide are given at~\cite{ta_ucc_nodate};
different from other code obfuscation implementations, \emph{uCc} uses cranelift~\cite{noauthor_cranelift_nodate}
as backend.

Our effort is to attack the core of type reconstruction algorithms,
these are the data flow used to build type constraints, and the \emph{semantic gap} between types in the high-level language and their
representations in the low-level machine code.
Beside custom tricks, \emph{uCc} uses extensively
obfuscation transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To make some code diversity, the compiler is \emph{probabilistic}: given a source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

\subsection*{Evaluation}
We test outputs of \emph{uCc} against some decompilers, table~\ref{description} describes briefly
tested functions and their types in source codes, table~\ref{flattening} shows results of
decompilers on binaries generated by \emph{uCc}. The effect of obfuscation is to make the types
of functions harder to be reconstructed. Indeed, the decompilers know very little about the original
types. Basically, any argument in the function signature is simply assigned by either \texttt{int64}
or \texttt{uint64}, the highest types in the \emph{conversion rank} of integer types, we call this
effect \emph{type flattening}.

\begin{table}
	\begin{center}
		\caption{Functions and original types}\label{description}
		\begin{tabular}{|l|l|}
			\hline
			Description & Function type \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{char mod(short, char)} \\
			\hline
			increase $p \leftarrow p + 1$ then dereference & \texttt{int inc\_deref(int *p)} \\
			\hline
			strlen  & \texttt{int slen(char* s)} \\
			\hline
			sdbm hash & \texttt{int sdbm(char *s)} \\
			\hline
			djb2 hash & \texttt{int djb2(char *s)} \\
			\hline
			sum of array $a$ & \texttt{long sum(short *a, int n)} \\
			\hline
			$n$-th fibonaci number (recursive impl.) & \texttt{int fibo(int n)} \\
			\hline
			num. of steps until $n \rightsquigarrow 1$ (Collatz's conj.)  & \texttt{int collatz(int n)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

@@

It may worth to note that the \emph{type flattening} is given by \emph{uCc}, not by the source code
complexity of tested functions. Indeed if they are compiled by, \emph{clang} or \emph{gcc} for example,
then their signatures are reconstructed precisely
by the decompilers. Last but not least, \emph{uCc} attacks directly on type reconstruction algorithms,
it doesn't use obfuscation techniques
% , for example making disassembling or function boundary detection hard,
whose side effects affect the type
reconstruction. Indeed, functions in tested binaries are perfectly detectable, disassemblable and
decompilable; only their decompiled types are in question.

>>>>

% \usepackage{listings}
% \lstset{basicstyle=\small\ttfamily,showspaces=false,frame=tb}
% \lstloadlanguages{C,C++,[x86masm]Assembler}

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the "assumed" original language.
In a statically typed language like C, the compiler doesn't preserve the high-level
type information in the generated machine code,
% Because most of information about high-level types is lost in compilation,
% the machine code
then the \emph{type reconstruction} (or type recovery) from machine code requires
special techniques, a survey up until 2015 can be referenced in~\cite{caballero_type_2016}, more
recent research
includes~\cite{noonan_polymorphic_2016, maier_typeminer_2019, katz_estimating_2016, robbins_minx_2016}.

@@

\subsection*{Type flattening obfuscation}
Under our interest, the studies in binary code type reconstruction can be roughly divided into
two directions: one is based on semantics of programs, the other is not. We focus on the former,
and also on methods based on static analysis. The research in this vein starts with
Mycroft~\cite{mycroft_type-based_1999}, and improved
in~\cite{lee_tie_2011,noonan_polymorphic_2016,elwazeer_scalable_2013, robbins_minx_2016}.
One of the notable improvement is to propose some notion of \emph{subtyping} to formally estimate
the closeness of inferred types over the original types. For example, the subtyping
in~\cite{lee_tie_2011} is based on a lattice which simulates the
\emph{conversion rank} of C's integer types, e.g. \texttt{int16\_t} is subtype of \texttt{int64\_t}.
Basically, if the inferred type is a supertype of the original one
then the result, though may be still sound, is less accurate. We should note that sometimes,
the supertype is the best result possible.

In this work, we limit on primitive types and their derived pointer types. Our purpose
is to neutralize the subtyping relation: a type can only be inferred as its
supertype. Concretely, \texttt{char}, \texttt{short}, \texttt{int}, etc. and pointers will
be recognized as either \texttt{uint64\_t} or \texttt{int64\_t}. We call this technique
\emph{type flattening obfuscation}.
Also we focus only on obfuscating types of function signatures, local variables are weaker primitives
since the compiler can remove them, e.g. through optimization.

\section{Compiler for type flattening}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types.
% \emph{uCc}'s objective is to make reconstructing C types from machine codes harder.
The compiler is implemented in Rust, the source code and a brief user guide are given at~\cite{ta_ucc_nodate};
different from other code obfuscation implementations, it uses cranelift~\cite{noauthor_cranelift_nodate}
as backend.

% Our effort is to
\emph{uCc} attacks the core of type reconstruction algorithms,
these are the data flow used to build type constraints and the \emph{semantic gap} between types in the high-level language and their
representations in the low-level machine code.
Beside custom tricks, \emph{uCc} uses extensively
transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To make some code diversity, the compiler is \emph{probabilistic}: given a source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

\begin{table}
	\begin{center}
		\caption{Functions and original types}\label{description}
		\begin{tabular}{|l|l|}
			\hline
			Description & Signature \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{char mod(short, char)} \\
			\hline
			increase $p \leftarrow p + 1$ then dereference & \texttt{int inc\_deref(int *p)} \\
			\hline
			strlen  & \texttt{int slen(char* s)} \\
			\hline
			sdbm hash & \texttt{int sdbm(char *s)} \\
			\hline
			djb2 hash & \texttt{int djb2(char *s)} \\
			\hline
			sum of array $a$ & \texttt{long sum(short *a, int n)} \\
			\hline
			$n$-th fibonaci number (recursive impl.) & \texttt{int fibo(int n)} \\
			\hline
			num. of steps until $n \rightsquigarrow 1$ (Collatz's conj.)  & \texttt{int collatz(int n)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

\subsection*{Evaluation}
We test outputs of \emph{uCc} against some decompilers, table~\ref{description} describes briefly
tested functions and their signatures in source codes, table~\ref{flattening} shows results of
decompilers on binaries generated by \emph{uCc}.
% The effect of obfuscation is to make the types of functions harder to be reconstructed.
% Indeed, the decompilers know very little about the original
% types:
Basically, any argument in the function signature is simply assigned by either \texttt{int64}
or \texttt{uint64}.
% , the highest types in the \emph{conversion rank} of integer types, we call this
% effect \emph{type flattening}.

@@

It may worth to note that the \emph{type flattening} is given by \emph{uCc}, not by the source code
complexity of tested functions. Indeed if they are compiled by, \emph{clang} or \emph{gcc} for example,
then their signatures can be reconstructed precisely
by any tested decompiler. Last but not least, \emph{uCc}
% attacks directly on type reconstruction algorithms,
% it
doesn't use obfuscation techniques
% , for example making disassembling or function boundary detection hard,
whose side effects may affect the type
reconstruction: functions in tested binaries are perfectly detectable, disassemblable and
decompilable; only their inferred signatures are in question.
